---
title: "Data 205 fp"
author: "Ike Charistan D"
format: html
editor: visual
---

## Load Libraries

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(plotly)
library(leaflet)
library(sf)
library(tigris)
library(scales)
library(readr)
```

## Load Datasets

```{r}
# Dataset 1 = Maryland grants and loans
setwd("~/Desktop/Data 110")
grants <- read.csv("State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2024.csv")

# Dataset 2 = Average wage per job
setwd("~/Desktop/Data 205")
wages <- read.csv("Maryland_Average_Wage_Per_Job_(Current_Dollars)__2014-2024_20260414.csv")

# Dataset 3 = ZIP code lookup table
zip_lookup <- read_csv("Zip_Code_Lookup_Table_20260414.csv")
```

## Clean & Prepare Data

```{r}
# Clean column names
grants <- clean_names(grants)
wages <- clean_names(wages)
zip_lookup <- clean_names(zip_lookup)

# Convert ZIP to text
zip_lookup <- zip_lookup %>%
  mutate(zip_code = as.character(zip_code))
```

## Reshaping the Wage Data

```{r}
# Convert wage data into long format for easier merging

wages_long <- wages %>%
  rename(year = 2) %>%
  select(year, montgomery_county, prince_george_s_county) %>%
  pivot_longer(cols = -year,
               names_to = "county",
               values_to = "avg_wage") %>%
  mutate(
    county = case_when(
      county == "montgomery_county" ~ "Montgomery County",
      county == "prince_george_s_county" ~ "Prince George's County"
    ),
    year = as.numeric(year)
  )
```

## Clean Grants Data

```{r}
# Match grants to counties using ZIP codes
# Then total yearly funding for each county

grants_clean <- grants %>%
  mutate(
    zip_code = as.character(zip_code),
    fiscal_year = as.numeric(fiscal_year)
  ) %>%
  left_join(zip_lookup, by = "zip_code") %>%
  filter(county %in% c("Montgomery County", "Prince George's County")) %>%
  group_by(fiscal_year, county) %>%
  summarize(
    total_grant_amount = sum(amount, na.rm = TRUE),
    .groups = "drop"
  )
```

## Merge Datasets

```{r}
# Final dataset combines funding and wages

df <- grants_clean %>%
  inner_join(wages_long,
             by = c("fiscal_year" = "year",
                    "county" = "county"))

df <- df %>%
  mutate(avg_wage = parse_number(avg_wage))
```

## Visualization 1: Annual State Funding Comparison

```{r}
ggplot(df, aes(x = factor(fiscal_year),
               y = total_grant_amount,
               fill = county)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "Montgomery County" = "#003366",
    "Prince George's County" = "#C8102E")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Annual State Grant Totals",
    x = "Year",
    y = "Funding ($)",
    fill = "County"
  ) +
  theme_minimal()
```

This graph compares how much state funding each county received each year.

## Visualization 2: Funding vs Wages

```{r}
ggplot(df, aes(x = total_grant_amount,
               y = avg_wage,
               color = county)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Grant Funding vs Average Wage",
    x = "Funding",
    y = "Average Wage"
  ) +
  theme_light()
```

Upward trend lines suggest counties receiving more funding also had higher wages.

## Visualization 3: 2023 County Comparison

```{r}
df_2023 <- df %>%
  filter(fiscal_year == 2023)

ggplot(df_2023, aes(x = reorder(county, total_grant_amount), 
                   y = total_grant_amount, 
                   fill = county)) +
  
  geom_col(show.legend = FALSE, color = "black", linewidth = 0.3) +
  
  geom_text(aes(label = dollar(total_grant_amount, scale = 1e-6, suffix = "M")),
            hjust = -0.2, size = 5, fontface = "bold") +
  
  scale_fill_manual(values = c(
    "Montgomery County" = "#005A9C",
    "Prince George's County" = "#E21833"
  )) +
  
  coord_flip() +
  
  scale_y_continuous(
    labels = label_dollar(scale = 1e-6, suffix = "M"),
    expand = expansion(mult = c(0, 0.2))
  ) +
  
  labs(
    title = "Total State Grant Funding by County",
    subtitle = "Fiscal Year 2023 Comparison",
    x = "County Name", 
    y = "Total Funding (Millions USD)"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 16, face = "bold", color = "black"),
    axis.text.x = element_text(size = 14),
    axis.title = element_text(size = 16, face = "bold"),
    plot.title = element_text(size = 20, face = "bold", color = "#333333"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

## CORRELATION ANALYSIS

```{r}
df %>%
  group_by(county) %>%
  summarize(
    correlation = cor(total_grant_amount,
                      avg_wage,
                      use = "complete.obs")
  )
```

**Interpretation**

"While funding and wage growth are strongly correlated, this relationship is associative rather than causal. Montgomery County’s higher correlation suggests more efficient conversion of public investment into high-wage sectors. In Prince George’s County, a more moderate correlation indicates that external factors—such as labor market conditions or industry mix—may dampen the economic impact of state funding."

## STATISTICAL TESTING:Pearson Correlation Test

**Q:** Is there statistically significant evidence of a positive linear relationship between state grant funding and average wages in each county?

$H_0$: p = 0 $H_a$: p \> 0

**Montgomery County**

```{r}
df %>%
  filter(county == "Montgomery County") %>%
  with(cor.test(total_grant_amount, avg_wage))
```


**Prince George's County**

```{r}
df %>%
  filter(county == "Prince George's County") %>%
  with(cor.test(total_grant_amount, avg_wage))
```


## PAIRED T-TEST

**Hypotheses**

$H_0$: $\mu_d$ = 0 $H_a$: $\mu_d$ $\neq$ 0

```{r}
mont <- df %>%
  filter(county == "Montgomery County") %>%
  arrange(fiscal_year) %>%
  pull(avg_wage)

pg <- df %>%
  filter(county == "Prince George's County") %>%
  arrange(fiscal_year) %>%
  pull(avg_wage)

t.test(mont, pg, paired = TRUE)
```
```{r}
paired_df <- df %>%
  select(fiscal_year, county, avg_wage) %>%
  pivot_wider(names_from = county,
              values_from = avg_wage) %>%
  mutate(diff =
           `Montgomery County` -
           `Prince George's County`)

ggplot(paired_df, aes(x = fiscal_year, y = diff)) +
  geom_col(fill = "#003366") +
  labs(
    title = "Annual Wage Gap",
    x = "Year",
    y = "Difference ($)"
  ) +
  theme_minimal()
```
